Kinematics \& Chemistry of Halo Substructures: The Vicinity of the Virgo Over-Density

### 2012-05-28 2011-sgr/aastex.tex
Initial commit.

\shorttitle{Kinematics \& Chemistry of Halo Substructures: The Vicinity of the VOD}
\shortauthors{Casey, Keller \& Da Costa}

@@

\author{Andrew R. Casey, Stefan Keller and Gary Da Costa}
\affil{Research School of Astronomy \& Astrophysics,
	Australian National University, Mt Stromlo Observatory, via Cotter Rd, Weston, ACT 2611, Australia}

\begin{abstract}
We present observations obtained with the AAT's 2dF wide field spectrograph AAOmega of K-type stars located within a region of the sky which contains the Virgo Over-Density and the leading arm of the Sagittarius Stream.  On the basis of the resulting velocity histogram we isolate populations in these overlapping regions, including Sagittarius and previously discovered Virgo groups. Through comparisons with \textit{N}-body models of the Galaxy-Sagittarius interaction, we find a tri-axial dark matter halo is favoured and we exclude a prolate shape. This result is contradictory with other observations along the Sagittarius leading arm, which typically favour prolate models.  We have also uncovered K-giant members of Sagittarius that are notably more metal poor ($\langle[$Fe/H$]\rangle = -1.7\,\pm\,0.3$) than previous studies. This suggests a significantly wider metallicity distribution exists in the Sagittarius Stream than formerly considered. We also present data on five serendipitous carbon stars which were uncovered in our sample.
\end{abstract}

@@

The proportion of substructure recently uncovered in the Galaxy has highlighted the crucial involvement accretion has played in the formation of the Milky Way \citep{Bell;et-al_2008}. Properties of these stellar structures allow us to probe the formation mechanisms and history of the Galaxy. Recent studies \citep{Carollo;et-al_2007, Carollo;et-al_2010} have suggested multiple evolution methods are required for galaxy formation to reconcile observational evidence, although this is a subject of ongoing debate \citep{Schoenrich;et-al_2010}. Regardless, the dissipation-less merging paradigm is widely accepted, and consistent with favoured Cold Dark Matter ($\Lambda$CDM) cosmology models. Through the examination of ongoing accretion events in the Milky Way and fossils from previous mergers, we can trace the evolution of the outer most regions of the Galaxy \citep[e.g.,][]{Helmi;White_2001}.

 Accretion is at least partly \citep[e.g.,][]{Starkenburg;et-al_2009}, if not entirely responsible for the formation of the stellar halo. \citet{Bell;et-al_2008} compared Sloan Digital Sky Survey \citep[hereafter SDSS]{York;et-al_2000} data to galaxy formation simulations using different dark halos and found that observations are consistent with the stellar halo being entirely formed by hierarchical merging of accreted satellites (see also \citet{Xue;et-al_2011}).

Unquestionably the most prominent ongoing accretion event within the Milky Way is that of the Sagittarius (Sgr) dwarf Spheroidal (dSph) galaxy. Originally uncovered by \citet{Ibata;et-al_1994} as a co-moving group of K- and M-type giants, the tidal tails of Sgr circle our Galaxy. As such they have been extensively traced with red-clump stars \citep{Majewski;et-al_1999}, carbon stars \citep{Totten;Irwin_1998, Ibata;et-al_2001}, RR Lyrae stars \citep{Ivezic;et-al_2000, Vivas;et-al_2005, Keller;et-al_2008, Watkins;et-al_2009, Prior;et-al_2009b}, A-type stars \citep{Newberg;et-al_2003}, BHB stars \citep{Ruhland;et-al_2011} and K/M-giants \citep{Majewski;et-al_2003,  Yanny;et-al_2009, Keller;Yong;Da_Costa_2010}. With spatial and kinematical information, tracers originating from the host system can be unequivocally identified. This is because they remain dynamically cold, and are identifiable as kinematic substructures long after they are stripped from their progenitor \citep[for example][]{Ibata;Lewis_1998, Helmi;White_1999}.

@@

\citet{Belokurov;et-al_2006} found an apparent bifurcation within the Sgr debris, which \citet{Fellhauer;et-al_2006} argued can only result from a dark halo having a near spherical shape. \citet{Law;et-al_2009} introduced a tri-axial model with a varying flattening profile $q$, which replicates the orbital precession seen and matched kinematic observations of the Sgr debris. However \citet{Law;Majewski_2010} (hereafter LM10) conceded this may be a purely numerical solution as tri-axial halos are dynamically unstable, and emphasizes more kinematic measurements in other regions of the Sgr stream are required.

@@

@@

When the presence of a stellar substructure is uncovered, K-giants provide excellent candidates for spectroscopic follow-up. They allow for precise radial velocities and chemical abundances. In order to specifically target K-giants, we have chosen candidates within the color selection box shown in Figure \ref{fig:cmd-target-selection}, taken from the SDSS DR7 catalogue. Field dwarfs are expected to contaminate the sample due to their similarity in colors. Although K-dwarfs are difficult to distinguish photometrically, we can spectroscopically separate these through the equivalent width (EW) of the  gravity-sensitive triplet Mg I lines at 5167.3, 5172.7, and 5183.6 \AA\ (see \S\ref{sec:dwarf-giant-separation}).

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./cmd_selection.eps}
	\caption{CMD of our observed regions from SDSS DR7, overlaid with the color selection criterion used to target K-giants. Appropriate Dartmouth isochrones \citep{Dotter;et-al_2008} are shown for Sgr debris at a distance of $r_\odot =40$ kpc \citep{Belokurov;et-al_2006}}
	\label{fig:cmd-target-selection}
\end{figure}

@@

Our targets were observed over two runs using AAOmega on the 3.9-m Anglo-Australian Telescope at Siding Springs Observatory in New South Wales, Australia. AAOmega is a double-beam, multi-object (392) fibre-fed spectrograph covering a two degree field of view. The targets were observed in normal visitor mode in April 2009. Throughout all observations, sufficient sky and guide fibres ($\sim30$ and 7-8 respectively) were allocated to ensure optimal sky subtraction and astrometry. In total 3,453 science targets were observed within 4 fields throughout the VOD/VSS region, as outlined in Table \ref{tab:observations}. Multiple configurations were observed for most fields to permit measurements of bright ($i < 16$) and faint ($i > 16$) stars, as well as repeat observations on a subset of stars.

% NOTE TO THE EDITOR - This table ought to be in a single column.

\begin{deluxetable}{cccc}
\tablecolumns{1}
\tablewidth{0pt}
\tabletypesize{\scriptsize}
\tablecaption{Observed fields\label{tab:observations}}
\tablehead{
	\colhead{Field} &
	\colhead{$\alpha$-center} &
	\colhead{$\delta$-center} &
	\colhead{Field} \\ 
	& (J2000.0) & (J2000.0) & Configurations
}
\startdata
A & 12 00 00 & $+00$ 00 00 & 1 \\ 
B & 12 20 00 & $-01$ 00 00 & 2 \\ 
C & 12 40 00 & $-02$ 00 00 & 3 \\ 
D & 12 56 00 & $-02$ 42 00 & 6 \\ 
\enddata
\end{deluxetable}

The beam was split into the red and blue arms using the 5700 \AA{} dichroic. The 580V grating in the blue arm yields spectra between 370-580 nm, with a resolution of $R = 1300$. In the red arm we used the 1000I grating ($R = 4400$) which spans the spectral range from 800-950 nm. This coverage includes the Ca II NIR Triplet (CaT), which is used for radial velocities and metallicities. To minimise scattered-light cross talk between fibres, science targets on each configuration were limited to 1.5 magnitudes in range. Globular clusters NGC 5024, 5053 and 5904 were observed as radial velocity and metallicity standards.

The data was reduced using the 2\textsc{DFDR}\footnote{http://www.aao.gov.au/AAO/2df/aaomega} pipeline. After being flat-fielded, the fibres were throughput calibrated and the sky spectrum was subtracted using the median of the dedicated sky fibres. Wavelength calibration was achieved from arc lamp exposures taken between each set of science fields. Multiple object frames were successively taken to assist with cosmic ray removal.

@@

A grid of synthetic spectra has been generated to quantitively establish a giant/dwarf star separation criterion. The grid was generated using \citet{Castelli;Kurucz_2004} model atmospheres with MOOG\footnote{http://www.as.utexas.edu/$\sim$chris/moog.html} and the line list of \citet{Kurucz;Bell_1995}. Spectra was also generated using stellar parameters for the Sun and Arcturus. The strength of the Mg I lines were tuned to match both the Solar and Arcturus atlases of \citet{Hinkle;et-al_2003}. \citet{Girardi;et-al_2004} isochrones have been used to translate our de-reddened $g - i$ color range to effective temperature. Reddening is accounted for using the maps of \citet{Schlegel;Finkbeiner;Davis_1998}. The corresponding effective temperature region ranges from 3900 to 5200 K, and is stepped at 25 K intervals. We have assumed typical K-type surface gravities of $\log{g} = 2$ for giants and $\log{g} = 4.5$ for dwarfs. Metallicities of [Fe/H] $= -0.5$, $-1.5$, and $-2.5$ were considered for both surface gravities.

All synthetic spectra was mapped onto the same wavelength space as our observations. Intensity was convolved with a Gaussian kernel of 3.03 \AA{} to match our observed resolution with the 580V grating. Mg I line strengths for our observations and synthetic spectra is shown against $g - i$ in Figure \ref{fig:dwarf-giant-separation}. As expected, there is an overlap of Mg I strengths between metal-rich ``giants'' ($\log{g} = 2$) and metal-poor ``dwarfs'' ($\log{g} = 4.5$). However, we do not expect metal-poor dwarfs to be a principle contaminant due to their intrinsically low luminosities and paucity.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./dwarf_giant_separation.eps}
	\caption{The sum of the Mg I triplet EWs for observations and synthetic spectra shown against the Sloan $g\,-\,i$ color. Synthetic spectra is shown for $\log{g} = 2$ ($\bigtriangleup$) and $\log{g} = 4.5$ ($\bigtriangledown$), and points are colored by metallicity. Our dwarf/giant separation line is shown (solid). }
	\label{fig:dwarf-giant-separation}
\end{figure}

 Using our synthetic grid, we have varied the slope and offset of our separation line to assess the effectiveness in differentiating giants from dwarfs.  A linear rule has been adopted to differentiate the populations, which maximises giant recoverability with minimal contamination. Using the rule,
\begin{equation}
	EW_{Mg\,I} < 3(g-i) + 0.5
\end{equation}
\noindent we identify 185 giants in our observed fields. An analysis of our 185 giant candidates revealed three stars with high proper motions, all of which lay close to our dwarf/giant separation line. This suggests they are dwarfs, and have been excluded. Some possible giant targets were discarded due to insufficient S/N, or because the Mg I triplet fell on bad columns of the detector. All efforts were made to minimise these losses. The distilled giant sample size is 178.

\section{Radial Velocities}
\label{sec:kinematics}

The Ca II triplet absorption lines at 8498.0, 8542.1 and 8662.1 \AA{}  have been used to measure radial velocities. These lines are strong, and easily identifiable in Red Giant Branch (RGB) stars even at low resolution.  Observed spectra have been cross-correlated with typical synthetic spectra of a K-giant ($T_{eff} = 4500$ K, $\log{g} = 2$, [Fe/H] $= -1.5$), and heliocentric corrections were made. Radial velocity measurements made on the standard stars in our globular clusters agree (within 3 km s$^{-1}$) with the catalogue of \citet{Harris_1996} (2011 edition). A number of our targets were observed on multiple fields, which allows us to calculate the internal measurement error. The differences between multiple measurements of the same target was calculated, and they form a half-normal distribution with a HWHM $= 3.58$ km s$^{-1}$.

@@

We have measured metallicities for our giants using the strength of the Ca triplet lines. This technique was originally empirically described for individual stars in globular  clusters \citep{Armandroff;Da-Costa_1991}. A spectroscopic analysis using VLT/FLAMES observations of RGB stars from composite populations led \citet{Battaglia;et-al_2008} to conclude that a calibrated CaT-[Fe/H] relationship can be confidently used in composite stellar populations \citep[see also][]{Rutledge;Hesser;Stetson_1997, Starkenburg;et-al_2010}. The caveat to this technique is that a luminosity (specifically $V - V_{HB}$) is required for calibration, and we have to assume a $V_{HB}$ luminosity here. Johnson V-band magnitudes were calculated from SDSS \textit{ugriz} photometry using \citet{Jester;et-al_2005} transformations. The weaker third Ca triplet line is more susceptible to noise and residual sky-line contamination \citep{Tolstoy;et-al_2001,Battaglia;et-al_2008}. Consequently, the strongest two CaT lines (8542 and 8662\AA) have been used to form a reduced EW (W') such that,

\begin{eqnarray}
	\textstyle\sum{W}\, &=& EW_{8542} + EW_{8662} \\
	W' &=&\textstyle\sum{W} + 0.64\left(\pm 0.02\right)\left(V-V_{HB}\right)
\end{eqnarray}
\noindent and the metallicity linearly varies with W' where,
\begin{equation}
	\mbox{[Fe/H]}_{\mbox{\small{CaT}}} = (-2.81\pm0.16) + (0.44\pm0.04) W'
	\label{eq:feh-cat}
\end{equation}

@@

@@

@@

@@

	\begin{figure}[h!]
		\includegraphics[width=\columnwidth]{./velocity_halo.eps}
		\caption{Generalised histogram of $V_{GSR}$ for our 178 K-type giants, highlighting their significance against a canonical halo population. The Gaussian is normalised such that the integral equals the number of observed stars excluding those outside a 2.5-$\sigma$ excess. Significant ($>3\sigma$) kinematic deviations from the smooth distribution are appropriately grouped and labelled.}
		\label{fig:velocity-histogram}
	\end{figure}

The generalised histogram represents each data point with a Gaussian kernel of an equal width (deviation). As our internal kinematic errors between multiple measurements summate to a half-normal distribution with a FWHM of 7.16 km s$^{-1}$, we have opted for a sigma value of 10 km s$^{-1}$ the generalised histogram in Figure \ref{fig:velocity-histogram} to avoid under-smoothing.

	\begin{figure}[h!]
		\includegraphics[width=\columnwidth]{./monte_carlo.eps}
		\caption{Monte-Carlo simulation results illustrating the number of times simulations could reproduce our observed data in the equivalent multi-dimensional bin. Significant features discussed in the text are labelled.}
		\label{fig:monte-carlo}
	\end{figure}

>>>>

\shorttitle{}
\shortauthors{Casey, Keller \& Da Costa}

@@

\author{Andrew R. Casey, Stefan Keller and Gary Da Costa}
\affil{Research School of Astronomy \& Astrophysics,
	Australian National University,
	Canberra, Australia}

\begin{abstract}
We present observations obtained with the AAT's 2dF wide field spectrograph AAOmega of K-type stars located within a region of the sky which contains the Virgo Over-Density and the leading arm of the Sagittarius Stream.  On the basis of the resulting velocity histogram we isolate populations in these overlapping regions, including Sagittarius and previously discovered Virgo groups. Through comparisons with \textit{N}-body models of the Galaxy-Sagittarius interaction, we find a tri-axial dark matter halo is favoured and we exclude a prolate shape. This result is contradictory with other observations along the Sagittarius leading arm, which typically favour prolate models.  We have also uncovered K-giant members of Sagittarius that are notably more metal poor ($\langle[$Fe/H$]\rangle = -1.7\,\pm\,0.3$) than previous studies. This suggests a significantly wider metallicity distribution exists in the Sagittarius Stream than formerly considered.
\end{abstract}

@@

The proportion of substructure recently uncovered in the Galaxy has highlighted the crucial involvement accretion has played in the formation of the Milky Way. Properties of these stellar structures allow us to probe the formation mechanisms and history of the Galaxy. Recent studies \citep{Carollo;et-al_2007, Carollo;et-al_2010} have suggested multiple evolution methods are required for galaxy formation to reconcile observational evidence, although this is a subject of ongoing debate \citep{Schoenrich;et-al_2010}. Regardless, the dissipation-less merging paradigm is widely accepted, and consistent with favoured Cold Dark Matter ($\Lambda$CDM) cosmology models. Through the examination of ongoing accretion events in the Milky Way and fossils from previous mergers, we can trace the evolution of the outer most regions of the Galaxy \citep[][e.g. ]{Helmi;White_2001}.

%ultimately trace the formation history of the Galaxy. 
%This technique of studying stellar populations as surrogates of galactic archaeology was first proposed in the seminal paper of \citet{Eggen;Lynden-Bell;Sandage_1962}, who observed a metallicity abundance declivity in stars from the disk to the halo. From these measurements they inferred that  the Galaxy formed by a rapid collapse of a relatively homogenous protogalactic cloud. In contrast of this inference, \citet{Searle;Zinn_1978} observed a wide scope of metallicities in globular clusters irrespective of their galactocentric distance. This implied a formation paradigm based on the hierarchical accretion of protogalactic fragments.

 Accretion is at least partly \citep[][e.g.]{Starkenburg;et-al_2009}, if not entirely responsible for the formation of the stellar halo. \citet{Bell;et-al_2008} compared Sloan Digital Sky Survey \citep[hereafter SDSS]{York;et-al_2000} data to galaxy formation simulations using different dark halos and found that observations are consistent with the stellar halo being entirely formed by hierarchical merging of accreted satellites (see also \citet{Xue;et-al_2011}).

Unquestionably the most prominent ongoing accretion event within the Milky Way is that of the Sagittarius (Sgr) dwarf Spheroidal (dSph) galaxy. Originally uncovered by \citet{Ibata;et-al_1994} as a co-moving group of K- and M-type giants, the tidal tails of Sgr circle our Galaxy. As such they have been extensively traced with red-clump stars \citep{Majewski;et-al_1999}, carbon stars \citep{Totten;Irwin_1998, Ibata;et-al_2001}, RR Lyrae stars \citep{Ivezic;et-al_2000, Vivas;et-al_2005, Keller;et-al_2008, Watkins;et-al_2009, Prior;et-al_2009b}, A-type stars \citep{Newberg;et-al_2003} and K/M-giants \citep{Majewski;et-al_2003,  Yanny;et-al_2009, Keller;Yong;Da_Costa_2010}. With spatial and kinematical information, tracers originating from the host system can be unequivocally identified. This is because they remain dynamically cold, and are identifiable as kinematic substructures long after they are stripped from their progenitor \citep[for example][]{Ibata;Lewis_1998, Helmi;White_1999}.

@@

\citet{Belokurov;et-al_2006} found an apparent bifurcation within the Sgr debris, which \citet{Fellhauer;et-al_2006} argued can only result from a dark halo having a near spherical shape. \citet{Law;et-al_2009} introduced a tri-axial model with a varying flattening profile $q$, which replicates the orbital precession seen and matched kinematic observations of the Sgr debris. However \citet{Law;Majewski_2010} (hereafter LM10) conceded this may be a purely numerical solution as tri-axial halos are dynamically unstable, and admits more kinematic measurements in other regions of the Sgr stream are required.

@@

% At the core of the VOD, multiple studies find a peak density of RR Lyrae \citep{Vivas;et-al_2001, Ivezic;et-al_2005} and F-type main sequence stars \citep{Newberg;et-al_2002} at ($\alpha, \delta$) = (190$^\circ$, 0$^\circ$). Close by, \citet{Juric;et-al_2008} find the F-type main sequence peak at ($\alpha, \delta$) = (192$^\circ$, +2$^\circ$) and \citet{Keller_2010} finds the sub-giant peak density at ($\alpha, \delta$) = (194.6$^\circ$, +1.8$^\circ$). The exact spatial centroid of the VOD remains equivocal.

%Furthermore, the spatial bounds of the VOD remains unclear. There is some agreement that the boundaries of the structure extend outside the SDSS catalogue \citep{Newberg;et-al_2007,Prior;et-al_2009a,Keller;et-al_2009}. Observations to date represent lower limits on the VOD bounds. \citet{Juric;et-al_2008} used photometric parallaxes to ascertain a spatial limit, and found the diffuse structure spanned $\sim1000$ deg$^{2}$. \citet{Prior;et-al_2009a} estimated a lower coverage of $\sim760$ deg$^2$ from the luminosity function excess in the region. \citet{Duffau;et-al_2006} had previously used a similar technique to \citet{Prior;et-al_2009a} and yielded a minimum sky coverage of 106 deg$^2$.

@@

When the presence of a stellar substructure is uncovered, K-giants provide excellent candidates for spectroscopic follow-up. They allow for precise radial velocities and chemical abundances. In order to specifically target K-giants, we have chosen candidates within the colour selection box shown in Figure \ref{fig:cmd-target-selection}, taken from the SDSS DR7 catalogue. Field dwarfs are expected to contaminate the sample due to their similarity in colours. Although K-dwarfs are difficult to distinguish photometrically, we can spectroscopically separate these through the equivalent width (EW) of the  gravity-sensitive triplet Mg I lines at 5167.3, 5172.7, and 5183.6 \AA\ (see \S\ref{sec:dwarf-giant-separation}).

%\begin{eqnarray}
%		0.6 <& (g-i)_o &< 1.7 \nonumber \\
%		15 <& i_o &< 18 \\
%-15(g-i)_o + 27 <& g_o &< -3.75(g-i)_o \nonumber
%\end{eqnarray}
%These criteria are marked in the CMD for this region in Figure \ref{fig:cmd-target-selection}. 
\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/cmd.eps}
	\caption{CMD of our observed regions from SDSS DR7, overlaid with the colour selection criterion used to target K-giants. Appropriate Dartmouth isochrones \citep{Dotter;et-al_2008} are shown for Sgr debris at a distance of $r_\odot =40$ kpc \citep{Belokurov;et-al_2006}}
	\label{fig:cmd-target-selection}
\end{figure}

@@

Our targets were observed over two runs using AAOmega on the 3.9-m Anglo-Australian Telescope at Siding Springs Observatory in New South Wales, Australia. AAOmega is a double-beam, multi-object (392) fibre-fed spectrograph covering a two degree field of view. The targets were observed in normal visitor mode in April 2009. Throughout all observations, sufficient sky and guide fibres ($\sim30$ and 7-8 respectively) were allocated to ensure optimal sky subtraction and astrometry. In total 3,453 science targets were observed across 4 fields (Table \ref{tab:observations}). Multiple configurations were observed for most fields to permit measurements of bright ($i < 16$) and faint ($i > 16$) stars, as well as repeat observations on a subset of stars.

\begin{deluxetable}{cccc}
\tablecolumns{1}
\tablewidth{0pt}
\tabletypesize{\scriptsize}
\tablecaption{Observed fields\label{tab:observations}}
\tablehead{
	\colhead{Field} &
	\colhead{$\alpha$-center} &
	\colhead{$\delta$-center} &
	\colhead{Field} \\ 
	& (J2000.0) & (J2000.0) & Configurations
}
\startdata
A & 12 00 00 & $+00$ 00 00 & 1 \\ % & 188 \\ %180
B & 12 20 00 & $-01$ 00 00 & 2 \\ %& 673 \\ %185
C & 12 40 00 & $-02$ 00 00 & 3 \\ %& 870 \\ %190
D & 12 56 00 & $-02$ 42 00 & 6 \\ %& 1,722 %194
\enddata
\end{deluxetable}

The beam was split into the red and blue arms using the 5700 \AA{} dichroic. The 580V grating in the blue arm yields spectra between 370-580 nm, with a resolution of $R = 1300$. In the red arm we used the 1000I grating which spans the spectral range from 800-950 nm. This coverage includes the Ca II NIR Triplet (CaT), which is used for radial velocities and metallicities. To minimise scattered-light cross talk between fibres, science targets on each configuration were limited to 1.5 magnitudes in range. Globular clusters NGC 5024, 5053 and 5904 were observed as radial velocity and metallicity standards.

The data was reduced using the 2\textsc{DFDR}\footnote{http://www.aao.gov.au/AAO/2df/aaomega} pipeline. After being flat-fielded, the fibres were throughput calibrated and the sky spectrum was subtracted using the median of the dedicated sky fibres. Wavelength calibration was achieved from arc lamp exposures taken between each set of science fields. Multiple object frames were successively taken to assist with cosmic ray removal. %Three thirty-minute science exposures were taken for each faint field, and three twenty-five minute exposures for bright fields.

@@

%\begin{figure*}[t]
%	\includegraphics[width=\textwidth]{./figures/spectra.eps}
%	\caption{Blue arm spectra for a K-type giant (top) and dwarf (bottom) star, highlighting the difference in the gravity-sensitive Mg I triplet profiles at 5167.3, 5172.7 and 5183.6 \AA{}, and the typical fitting achieved.}
%	\label{fig:dwarf-giant-comparison}
%\end{figure*}

A grid of synthetic spectra has been generated to quantitively establish a giant/dwarf star separation criterion. The grid was generated using \citet{Castelli;Kurucz_2004} model atmospheres with MOOG\footnote{http://www.as.utexas.edu/$\sim$chris/moog.html} and the line list of \citet{Kurucz;Bell_1995}. Spectra was also generated using stellar parameters for the Sun and Arcturus. The strength of the Mg I lines were tuned to match both the Solar and Arcturus atlases of \citet{Hinkle;et-al_2003}. \citet{Girardi;et-al_2004} isochrones have been used to translate our de-reddened $g - i$ colour range to effective temperature. Reddening is accounted for using the maps of \citet{Schlegel;Finkbeiner;Davis_1998}. The corresponding effective temperature region ranges from 3900 to 5200 K, and is stepped at 25 K intervals. We have assumed typical K-type surface gravities of $\log{g} = 2$ for giants and $\log{g} = 4.5$ for dwarfs. Metallicities of [Fe/H] $= -0.5$, $-1.5$, and $-2.5$ were considered for both surface gravities.

All synthetic spectra was mapped onto the same wavelength space as our observations. Intensity was convolved with a Gaussian kernel of 3.03 \AA{} to match our observed resolution with the 580V grating. Mg I line strengths for our observations and synthetic spectra is shown against $g - i$ in Figure \ref{fig:dwarf-giant-separation}. As expected, there is an overlap of Mg I strengths between metal-rich ``giants'' ($\log{g} = 2$) and metal-poor ``dwarfs'' ($\log{g} = 4.5$). However, we do not expect metal-poor dwarfs to be a principle contaminant due to their intrinsically low luminosities and paucity.
\begin{figure}[h]
	\includegraphics[width=\columnwidth]{./figures/dwarfgiants.eps}
	\caption{The sum of the Mg I triplet EWs for observations and synthetic spectra shown against the Sloan $g\,-\,i$ colour. Synthetic spectra is shown for $\log{g} = 2$ ($\bigtriangleup$) and $\log{g} = 4.5$ ($\bigtriangledown$), and points are coloured by metallicity. Our dwarf/giant separation line is shown (solid). }
	\label{fig:dwarf-giant-separation}
\end{figure}
 %The choice of separation criterion affects our dwarf contamination, and our giant recoverability. 
 Using our synthetic grid, we have varied the slope and offset of our separation line to assess the effectiveness in differentiating giants from dwarfs.  A linear rule has been adopted to differentiate the populations, which maximises giant recoverability with minimal contamination. Using the rule,
\begin{equation}
	EW_{Mg\,I} < 3(g-i) + 0.5
\end{equation}
\noindent we identify 185 giants in our observed fields. An analysis of our 185 giant candidates revealed three stars with high proper motions, all of which lay close to our dwarf/giant separation line. This suggests they are dwarfs, and have been excluded. Some possible giant targets were discarded due to insufficient S/N, or because the Mg I triplet fell on bad columns of the detector. All efforts were made to minimise these losses. The distilled giant sample size is 178.

\section{Radial Velocities}
\label{sec:kinematics}
%Their prominence allows us to determine accurate radial velocities even in noisy data.

The Ca II triplet absorption lines at 8498.0, 8542.1 and 8662.1 \AA{}  have been used to measure radial velocities. These lines are strong, and easily identifiable in Red Giant Branch (RGB) stars even at low resolution.  Observed spectra have been cross-correlated with typical synthetic spectra of a K-giant ($T_{eff} = 4500$ K, $\log{g} = 2$, [Fe/H] $= -1.5$), and heliocentric corrections were made. Radial velocity measurements made on the standard stars in our globular clusters agree (within 3 km s$^{-1}$) with the catalogue of \citet{Harris_1996} (2011 edition). A number of our targets were observed on multiple fields, which allows us to calculate the internal measurement error. The differences between multiple measurements of the same target was calculated, and they form a half-normal distribution with a HWHM $= 3.58$ km s$^{-1}$. %This measurement includes velocity measurements of identified dwarfs, as we are focussed here on determining the systematic accuracy of our velocity measurements and not identifying substructures.

%\begin{figure}[h]
%	\includegraphics[width=\columnwidth]{./figures/vel_error.eps}
%	\caption{Internal radial velocity errors between multiple measurements. This sample includes all targets (dwarfs and giants) with sufficient S/N that were observed more than once.}
%	\label{fig:multiple-kinematics}
%\end{figure}

@@

We have ascertained metallicities for our giants using the strength of the Ca triplet lines. This technique was originally empirically described for individual stars in globular  clusters \citep{Armandroff;Da-Costa_1991}. A spectroscopic analysis using VLT/FLAMES observations of RGB stars from composite populations led \citet{Battaglia;et-al_2008} to conclude that a calibrated CaT-[Fe/H] relationship can be confidently used in composite stellar populations \citep[see also][]{Rutledge;Hesser;Stetson_1997, Starkenburg;et-al_2010}. The caveat to this technique is that a luminosity (specifically $V - V_{HB}$) is required for calibration, and we have to assume a $V_{HB}$ luminosity here. Johnson V-band magnitudes were calculated from SDSS \textit{ugriz} photometry using \citet{Jester;et-al_2005} transformations. The weaker third Ca triplet line is more susceptible to noise and residual sky-line contamination \citep{Tolstoy;et-al_2001,Battaglia;et-al_2008}. Consequently, the strongest two CaT lines (8542 and 8662\AA) have been used to form a reduced EW (W') such that,
\begin{comment}
%Nevertheless the CaT-[Fe/H] relationship is strong for RGB stars in composite populations within the range $-2.5 < \mbox{[Fe/H]} < -0.5$.

%However, a systematic trend in this high resolution spectra was found where metallicity is overestimated by $\sim0.1$ dex at low metallicities ($\mbox{[Fe/H]} < -2.2$) and underestimated by $\sim0.1-0.2$ dex at higher metallicities ($\mbox{[Fe/H]} > -1.2$) using this calibration. \citet{Battaglia;et-al_2008} indicate higher Ca abundances may affect the CaT equivalent width sufficiently to cause this trend, although [Fe/H] is the certain dominant factor contributing to the CaT line strength.

%The [Ca/H] - [Fe/H] association is linear, although this technique is only applicable for stars brighter than the horizontal branch ($V - V_{HB} > 0$). Transformations from SDSS $ugriz$ photometry to $UBVR_{C}I_{C}$ has been performed using the transformation equations in \citet{Jester;et-al_2005}. Photometric errors (including those from transformations) are propagated to our reported metallicity errors.

%The sum of the CaT EW is scaled to form a 'reduced EW' (W'). There is some discussion in which combination of the three CaT lines should be used. In this work we have taken the total equivalent width as the sum of the two stronger Ca lines ($\lambda_{8542}$, $\lambda_{8662}$). The weakest Ca line is usually the most unreliable when calculating equivalent widths due to noise and the possibility of residual sky-line contamination. Moreover, following a comparison of scaling relations between CaT and [Fe/H], \citet{Battaglia;et-al_2008} found using only the two strongest lines (as \citet{Tolstoy;et-al_2001} had done) was the most robust. As such, we have adopted the best-fitting relationship found by \citet{Battaglia;et-al_2008} for this work. The reduced EW is calculable by
\end{comment}
\begin{eqnarray}
	\textstyle\sum{W}\, &=& EW_{8542} + EW_{8662} \\
	W' &=&\textstyle\sum{W} + 0.64\left(\pm 0.02\right)\left(V-V_{HB}\right)
\end{eqnarray}
\noindent and the metallicity linearly varies with W' where,
\begin{equation}
	\mbox{[Fe/H]}_{\mbox{\small{CaT}}} = (-2.81\pm0.16) + (0.44\pm0.04) W'
	\label{eq:feh-cat}
\end{equation}

@@

\begin{comment}
%When applying this technique to our K-giants, the difficulty is that $V_{HB}$ is not easily defined. The horizontal branch magnitude varies between stellar systems, and is dependent on the heliocentric distance to, and morphology of, the stellar environment.

%Observations here include a composite population. Some stars are members of the Sgr leading arm, others are members of the VOD/VSS (see \S\ref{sec:substructure-identification}), all superimposed upon a background of halo stars. We can identify some substructures kinematically, but we cannot differentiate halo stars within those velocity bins. However in our observed region there are known dominant structures. The Sagittarius stream dominates our negative $V_{GSR}$ population, and our positive kinematic space is largely comprised of members from the VOD (see \S\ref{sec:substructure-identification}). As such we have separated our population at $V_{GSR} = 0$ into two samples to ascertain $V_{HB}$ for the two most dominant populations; Sgr and the VOD. Although this technique introduces a (known) systemic effect, it is the most useful way to estimate the metallicity distribution of each population.
\end{comment}

@@

\begin{comment}
The horizontal branch magnitude for the Sgr debris varies with age and distance. \citet{Belokurov;et-al_2006} determined distances along the two branches of the Sgr bifurcation using a colour-magnitude comparison with the Sgr main body \citep{Bellazzini;et-al_2006}. After being scaled for a revised distance modulus to the Sgr core \citep[$(m-M)_{0} = 17.27;$][]{Siegel;et-al_2007}, these distance measurements were used to constrain the more recent tri-axial halo simulation by LM10.  We can use these observed distances (as tabulated in Table 1 of LM10) and the known absolute magnitude of RR Lyrae stars \citep[$M_V = +0.69;$][]{Tsujimoto;et-al_1998} to calculate the horizontal branch magnitude for each position of the Sgr stream. Reddening is accounted for each star using the maps of \citet{Schlegel;Finkbeiner;Davis_1998} and is scaled to the V-band with $A/E(B-V) = 3.240$. A spline is fitted to the tabulated right ascension and distance measurements, and the distance to each star is found by evaluating the right ascension of each star along the spline. Distance to the branch does not vary significantly with declination. The distance error is taken as the largest error from the two neighbouring distance measurements.
%Distance errors are also propagated through to our reported metallicity error, which is at most $\sim 0.16$ dex. 
\end{comment}

@@

		%Along our line of sight, the spatial region examined in this paper incorporates the superposition of multiple substructures, some of which may be inter-related. To facilitate a logical flow of discussion, these substructures are identified first and discussed separately.

@@

	\begin{figure}[h!]
		\includegraphics[width=\columnwidth]{./figures/velocity-halo.eps}
		\caption{Generalised histogram of $V_{GSR}$ for our 178 K-type giants, highlighting their significance against a canonical halo population. The Gaussian is normalised such that the integral equals the number of observed stars excluding those outside a 2.5-$\sigma$ excess. Significant ($>3\sigma$) kinematic deviations from the smooth distribution are appropriately grouped and labelled.}
		\label{fig:velocity-histogram}
	\end{figure}

The generalised histogram represents each data point with a Gaussian kernel of an equal bandwidth (deviation). As our internal kinematic errors between multiple measurements summate to a half-normal distribution with a FWHM of 7.16 km s$^{-1}$, we have opted for a bandwidth value of 10 km s$^{-1}$ the generalised histogram in Figure \ref{fig:velocity-histogram} to avoid under-smoothing.

	\begin{figure}[h!]
		\includegraphics[width=\columnwidth]{./figures/montecarlo.eps}
		\caption{Monte-Carlo simulation results illustrating the number of times simulations could reproduce our observed data in the equivalent multi-dimensional bin. Significant features discussed in the text are labelled.}
		%\caption{Results from our 10,000 binned kinematic and chemical Monte-Carlo simulations, overlaid with our observed K-giant sample properties. Each bin is coloured by the number of times Monte-Carlo simulations reproduced at least the same number of observed stars in that bin. Red bins indicate particularly rare clumps, and blue bins demonstrate common samples.}
		\label{fig:monte-carlo}
	\end{figure}
%We have looked at stars in our dataset with repeated observations to gain an approximation for the bandwidth selection.
% Their statistical significance is determined by a comparison to a canonical halo population \citep[$\mu = 0$ km s$^{-1}, \sigma = 101.6$ km s$^{-1}$][]{Sirko;et-al_2004}.
